Choosing Color Palettes for Statistical Graphics

### 2006-10-23 trunk/vcd/inst/doc/hcl-colors.Rnw
started vignette for HCL palettes, removed jss.bst and vcd.bib (are assumed to be in texmf)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@601 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

@@

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. Usually, these should give the same
perceptual weight to each category so that no group is perceived to be larger/more
important than any other one. Typical applications of qualitative palettes in
statistics would be bar plots, pie charts (see Figure~\ref{fig:seats}) 
or highlighted mosaic displays (see Figure~\ref{fig:votes}).

@@

Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admissible values of in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-s-construct} were chosen because they correspond to similar
geometric shapes in the chroma/luminance plane, allowing for both large chroma and
luminance contrasts. If potential viewers of the resulting graphic might be color-blind,
the pair of hues should be taken from the yellow/blue axis of the color wheel
rather than the green/red axis as contrasts on the latter axis are more difficult to 
distinguish for color-blind people.

@@

In this section, we show a collection of examples for the various types of palettes applied
to statistical graphics. The first examples illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''
are employed. In this election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$ for all hues
leading to a much better balanced palette compared to the fully saturated
colors usually found in the (German) media (which instead of blue typically use the extreme
``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

@@

To illustrate sequential palettes, a bivariate density estimation for the Old
Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990} is visualized.
The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
and it is of some interest to understand the relation ship between the duration
of a geyser eruption and the waiting time until the next eruption.
Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
duration and waiting time revealing a multi-modal bivariate distribution with more
variation in the waiting time after a longer eruption. Both versions of the heatmap use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
chosen such that the peaks in the density are brought out very clearly by using a rather
large luminance contrast and $p_2 > 1$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde, height=6, width=12>>=
if(generate_png) {
  png(file = "bkde.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  data("geyser", package = "MASS")
  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  dev.off()
}
@
\includegraphics{bkde}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

@@

Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [30, 90]$)
and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
the previous example. For Figure~\ref{fig:class}, we set $p_1$ and $p_2$ to the somewhat
smaller value of $1.3$ to bring out all deviations from the decision boundary and not only the
extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
$p_1$ and $p_2$ is not so important because only 5 colors are selected---we use values
$p_1 = p_2 = 1$.

@@

Implementing the different color palettes suggested in the previous section 
is extremely easy if the software environment chosen already provides an implementation
of HCL colors: from the formulas provided above the HCL coordinates for a
palette can easily be computed. A bit more work is required if the sofware
package does not yet provide an HCL implementation. In that case, additional
functionality is needed for translating HCL coordinates to the software package's 
color system which may vary between different packages, but standardized RGB (sRGB)
is often used. The typical way of coordinate conversion is to go first from HCL to
CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
$U$ and $V$. Subsequently CIELUV is converted to CIEXYZ which in turn is converted
to sRGB (with the latter conversion depending on the device used for display).
The details of these conversions are somewhat technical and tedious (and hence
omitted here), however the conversion formulas are still straightfoward to
implement and can for example be found in \cite{vcd:Wiki+HSV:2006} or
\cite{vcd:Poynton:2000}.

@@

Qualitative palettes are implemented in \pkg{vcd} in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
where \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates. It computes the HCL coordinates
and transforms them to RGB coordinates by calling the \code{hcl()} function 
provided by the \pkg{colorspace} package.

@@

We provide paths through HCL space along perceptual axes
so that colors selected along these paths match perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:illustrations}, chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance. Another parameter in these 
palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
Employing a function $|i|^p$ for this, small
powers $p$ can be used if palettes with a lot of color should be 
constructed (e.g., to clearly separate regions in a classification map). On
the other hand, large powers $p$ will result in palettes with less color
(e.g., appropriate to highlight only extreme regions firmly classified in such a map).

>>>>

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006} and the HCL
color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

@@

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. Usually, these should give the same
perceptual weight to each category so that no group is perceived to be larger or more
important than any other one. Typical applications of qualitative palettes in
statistics would be bar plots \citep[see][]{vcd:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
or highlighted mosaic displays (see Figure~\ref{fig:votes}).

@@

Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admissible values in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-d-construct} were chosen because they correspond to similar
geometric shapes in the chroma/luminance plane, allowing for both large chroma and
luminance contrasts. If potential viewers of the resulting graphic might be color-blind,
the pair of hues should be taken from the yellow/blue axis of the color wheel
rather than the green/red axis as contrasts on the latter axis are more difficult to 
distinguish for color-blind people.

@@

In this section, we show a collection of examples for the various types of palettes applied
to statistical graphics. The first example illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''.
In this election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$ for all hues
leading to a much better balanced palette compared to the fully saturated
colors usually found in the (German) media (which instead of blue typically use the extreme
``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

@@

To illustrate sequential palettes, a bivariate density estimation for the Old
Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990} is visualized.
The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
and it is of some interest to understand the relation ship between the duration
of a geyser eruption and the waiting time for this eruption.
Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
duration and waiting time revealing a multi-modal bivariate distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes). Both versions of the heatmap use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
chosen such that the peaks in the density are brought out very clearly by using a rather
large luminance contrast and $p_2 > 1$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde, height=6, width=12>>=
if(generate_png) {
  png(file = "bkde.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  data("geyser", package = "MASS")
  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
  geyser2 <- cbind(geyser$duration, geyser$waiting)
  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  dev.off()
}
@
\includegraphics{bkde}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

@@

Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
the previous example. For Figure~\ref{fig:class}, we set $p_1$ and $p_2$ to the somewhat
smaller value of $1.3$ to bring out all deviations from the decision boundary and not only the
extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
$p_1 = p_2 = 1$.

@@

Implementing the different color palettes suggested in the previous section 
is extremely easy if the software environment chosen already provides an implementation
of HCL colors: from the formulas provided above the HCL coordinates for a
palette can be conveniently computed. A bit more work is required if the software
package does not yet provide an HCL implementation. In that case, additional
functionality is needed for translating HCL coordinates to the software package's 
color system which may vary between different packages, but standardized RGB (sRGB)
is often used. The typical way of coordinate conversion is to go first from HCL to
CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
$U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
to sRGB (with the latter conversion depending on the device used for display).
The details of these conversions are somewhat technical and tedious (and hence
omitted here), however the conversion formulas are still straightfoward to
implement and can, for example, be found in \cite{vcd:Wiki+LUV:2006} or
\cite{vcd:Poynton:2000}.

@@

Qualitative palettes are implemented in \pkg{vcd} in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
where \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates. It computes the HCL coordinates
and transforms them to RGB by calling the \code{hcl()} function 
provided by the \pkg{colorspace} package.

@@

We provide paths through HCL space along perceptual axes
so that colors selected along these paths match perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:illustrations}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance. Another parameter in these 
palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
Employing a function $|i|^p$ for this, small
powers $p$ can be used if palettes with a lot of color should be 
constructed (e.g., to clearly separate regions in a classification map). On
the other hand, large powers $p$ will result in palettes with less color
(e.g., appropriate to highlight only extreme regions firmly classified in such a map).
